Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-02-22 code/eda-code/eda.qmd
Part 2 Updates
Explored standardization methods to remove the effect of outliers before finding the population count variable `wra` which was used to standardize. Updated the processing file to include that variable and updated the manuscript with a more focused research question and the appropriate variables.

---
title: "An example exploratory analysis script"
author: "Andreas Handel"
date: "10/28/2021"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

@@

explore how many people use each method
explore how many countries are in each category for each method using new factor variable
explore percentage of unintended pregnancies across methods

# Data exploration through tables
We will begin with creating a summary table of our cleaned variables using the skim()function. This provides information

```{r}
summary_df = skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

@@

Percent of unintended pregnancies is the main outcome of interest, so we will explore this variable first.

```{r}
p1 <- mydata %>% ggplot(aes(x=pct_upreg)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

It is important to consider the type of birth control methods associated with percent of unintended pregnancies as well. We will create a histogram for the `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse`.

```{r}
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2)

p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3)

p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4)

p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

We will now explore the new 
plots 6-10

We will start exploring how many women use each type of birth control compared to the percent of unintended pregnancies related to that birth control. Plot 11 shows `pct_upreg_larcster` vs `total_larcster` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods.

```{r}
p11 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=total_larcster)) + geom_point() + geom_smooth(method='lm')
plot(p11)
figure_file = here("results","height_weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results","height_weight_stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

Our next group of plots will be violin plots to explore the usage level of birth associated with percentage of unintended pregnancies use for each birth control method. 
```{r}
p16 <- mydata %>% ggplot(aes(x=larcster_usage, y=pct_upreg_larcster))
```

Our last plots will be mosaic plots to compare two categorical variables: region and usage for each type of birth control. 
```{r}
library(ggmosaic)
p17 <- mydata %>% ggplot(aes(x=region, y=larcster_usage, fill = freq)) + geom_mosaic(aes(x=region, fill = larcster_usage))
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

---
title: "Exploratory Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "02/21/24"
output: html_document
---

@@

# Data exploration through tables
We will begin with creating a summary table of our cleaned variables using the skim()function. This provides information about each of the numeric variables including the mean, standard deviation, quartiles, missing observations, and a histogram.

```{r}
# skim the data to create summary file
summary_df = skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
We also created simple tables for the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}

```

@@

It is important to consider the type of birth control methods associated with percent of unintended pregnancies as well. We will create a histogram for the `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse` variables.

```{r}
# explore the percentage of unintended pregnancies related to long acting reversible methods in each country
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2)

# explore the percentage of unintended pregnancies related to short acting reversible methods in each country
p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3)

# explore the percentage of unintended pregnancies related to traditional methods in each country
p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4)

# explore the percentage of unintended pregnancies related to no birth control usage in each country
p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

We will now explore the new total usage variables that were created using the number of women using each method across the four age categories. This set of plots revealed large outliers in the 'total_larcster`, 'total_sarc', 'total_trad', and `total_none` variables, which prevented the graphs from being very useful. The histograms look like most observations for each of these four variables are 0, but that is not possible because the summary table showed that the minimum value for all four of these variables is greater than 0. This distortion is caused by the two large outliers in the data set.

```{r}
# explore the total number of people using long acting reversible methods in each country
p6 <- mydata %>% ggplot(aes(x=total_larcster)) + geom_histogram() 
plot(p6)
figure_file = here("results", "figures", "exploratory-figures", "total_larcster_distribution.png")
ggsave(filename = figure_file, plot=p6)

# explore the total number of people using short acting reversible methods in each country
p7 <- mydata %>% ggplot(aes(x=total_sarc)) + geom_histogram() 
plot(p7)
figure_file = here("results", "figures", "exploratory-figures", "total_sarc_distribution.png")
ggsave(filename = figure_file, plot=p7)

# explore the total number of people using traditional methods in each country
p8 <- mydata %>% ggplot(aes(x=total_trad)) + geom_histogram() 
plot(p8)
figure_file = here("results", "figures", "exploratory-figures", "total_trad_distribution.png")
ggsave(filename = figure_file, plot=p8)

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + geom_histogram() 
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```

We need to explore where the outliers in these variables are coming from. China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# find the countries that represent the outliers
outliers <- mydata %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

# standardization attempt 1
mydata <- mydata %>% mutate(larcster_st = total_larcster / sum(total_larcster),
                            sarc_st = total_sarc / sum(total_sarc),
                            trad_st = total_trad / sum(total_trad),
                            none_st = total_none / sum(total_none))

# standardization attempt 2
scaling_factor_larcster <- max(mydata$total_larcster)
mydata <- mydata %>% mutate(standardized_larcster = total_larcster / scaling_factor_larcster)

# standardization attempt 3
min_count <- min(mydata$total_larcster)
max_count <- max(mydata$total_larcster)
scaled_counts <- (mydata$total_larcster - min_count) / (max_count - min_count)
print(scaled_counts)
mean_count <- mean(mydata$total_larcster)
sd_count <- sd(mydata$total_larcster)
z_score_counts <- (mydata$total_larcster - mean_count) / sd_count
```

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reprodutive age. 
```{r}
# population size standarization for total counts per birth control method
mydata <- mydata %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```

We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + geom_point() + geom_smooth(method='lm')
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10)

# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
p11 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + geom_point() + geom_smooth(method='lm')
plot(p11)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=p11)

# explore percentage of unintended pregnancies compared to standardized count of women using traditional methods
p12 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard)) + geom_point() + geom_smooth(method='lm')
plot(p12)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totaltrad.png")
ggsave(filename = figure_file, plot=p12)

# explore percentage of unintended pregnancies compared to standardized count of women not using no method
p13 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard)) + geom_point() + geom_smooth(method='lm')
plot(p13)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalnone.png")
ggsave(filename = figure_file, plot=p13) 
```

We are also interested in how these relationships between percent of unintended pregnancies using each birth control method and the number of people using each method varies across region. We adjusted the original scatter plots created above by coloring based on region.

```{r}
# explore percent unintended pregnancies using long acting reversible and sterilziation methods by region
p14 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p14)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_stratified.png")
ggsave(filename = figure_file, plot=p14)

# explore percent unintended pregnancies using short acting reversible methods by region
p15 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p15)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png")
ggsave(filename = figure_file, plot=p15)

# explore percent unintended pregnancies using traditional methods by region
p16 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p16)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_stratified.png")
ggsave(filename = figure_file, plot=p16)

# explore percent unintended pregnancies using no method by region
p17 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p17)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_none_stratified.png")
ggsave(filename = figure_file, plot=p17) 
```

Our next group of plots will be violin plots to explore the usage level of each birth control method, a categorical variable, associated with percentage of unintended pregnancies use for each birth control method, a numerical variable.
```{r}
# explore long acting reversible and sterilization method usage compared to percent of unintended pregnancies
p18 <- mydata %>% ggplot(aes(x = larcster_usage, y = pct_upreg_larcster)) +
  geom_violin() +
  labs(x = "Long Acting Reversible and Sterilization Usage", y = "Percentage Unintended Pregnancies") 
plot(p18)
figure_file = here("results", "figures", "exploratory-figures", "larcster_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p18)

# explore short acting reversible method usage compared to percent of unintended pregnancies
p19 <- mydata %>% ggplot(aes(x = sarc_usage, y = pct_upreg_sarc)) +
  geom_violin() +
  labs(x = "Short Acting Reversible Usage", y = "Percentage Unintended Pregnancies") 
plot(p19)
figure_file = here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p19)

# explore traditional method usage compared to percent of unintended pregnancies
p20 <- mydata %>% ggplot(aes(x = trad_usage, y = pct_upreg_trad)) +
  geom_violin() +
  labs(x = "Traditional Method Usage", y = "Percentage Unintended Pregnancies") 
plot(p20)
figure_file = here("results", "figures", "exploratory-figures", "trad_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p20)

# explore no method usage compared to percent of unintended pregnancies
p21 <- mydata %>% ggplot(aes(x = none_usage, y = pct_upreg_nouse)) +
  geom_violin() +
  labs(x = "No Birth Control Usage", y = "Percentage Unintended Pregnancies") 
plot(p21)
figure_file = here("results", "figures", "exploratory-figures", "none_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p21) 
```

### 2024-02-22 products/manuscript/Manuscript.qmd
Part 2 Updates
Explored standardization methods to remove the effect of outliers before finding the population count variable `wra` which was used to standardize. Updated the processing file to include that variable and updated the manuscript with a more focused research question and the appropriate variables.

# Research Questions
1) What is the relationship between rates of unintended pregnancies and usage of contraceptive methods across various age groups?
The main outcome variables will be total percentage of unintended pregnancies (`pct_upreg`). 
The predictor variables will include percentage of unintended pregnancies across three types of birth control: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). No method variables can serve as control groups for statistical tests.
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

@@

2) How does marriage status affect percentage of unintended pregnancies across different types of birth control?
  -outcome variables:
      unintended pregnancies, total number (pct_upreg)
      unintended pregnancies, using larc and sterilization (pct_upreg_larcster)
      unintended pregnancies, using short term reversible (pct_upreg_sarc)
  -predictor variables:
      Number wanting to avoid pregnancy, all married women (inneed_married)
      Number using sterilization, all married women	(ster_married) **combine these two in processing
      Number using long-acting reversible methods, all married women	(larc_married)
      Number using short-acting methods, all married women	(sarc_married)
      Number wanting to avoid pregnancy, formerly married women 	(inneed_formerlymar)
      Number using sterilization, formerly married women	(ster_formerlymar)
      Number using long-acting reversible methods, formerly married women	(larc_formerlymar)
      Number using short-acting methods, formerly married women	(sarc_formerlymar)
      Number wanting to avoid pregnancy, never married women 	(inneed_nevermar)
      Number using all modern methods, never married women	(modmethods_nevermar)
      Number using sterilization, never married women	(ster_nevermar)
      Number using long-acting reversible methods, never married women	(larc_nevermar)
      Number using short-acting methods, never married women	(sarc_nevermar)

>>>>

# Research Questions
1) Is there a significant difference in the percentage of unplanned pregnancies between contraceptive methods among all countries? Does this difference vary by region?
The main outcome variables will be total percentage of unintended pregnancies per birth control method (`pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`). 
The control variable will be total percentage of unintended pregnancies among women not using birth control (`pct_upreg_none`)
The predictor variables will include the total number of people women using each type of birth control standardized by population size, which will be utilized to create a factor variable showing low, medium, and high usage for each method per country. 
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

@@
